---
title: "Housing Prices? How predictable!"
date: 'Summer 2019'
author: James Banasiak (jamesmb3), Ryan Epp (ryanepp2) and Nana Tark (ytark2)
version: 0.0.4
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***

# Introduction
In this data analysis project we explore the dataset from the *House Prices: Advanced Regression Techniques* competition held on [Kaggle](https://www.kaggle.com/c/house-prices-advanced-regression-techniques). The dataset describes residential homes in Ames, Iowa. The goal of is project (and the Kaggle competition) is to predict the sale price of a home based on it's other attributes. Theses attributes are represented in the dataset by an extensive 79 variables, including everything from the number of bathrooms, to the material the roof is made out of, to the year the home was built.

Our primary goal is to build a model which can predict house prices as accurately as possible. We want our model to work well with previously unseen examples. So to avoid overfitting, we will use cross validation accuracy to evaluate our candidate models.

As a secondary goal, we'd like our model to be stable and interperatable so we will also consider BIC, and Adjusted $R^2$ when selecting a best model.

***

# Methods

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
# library(boot)
# library(rsq)
# library(lmtest)
# library(knitr)
# library(forcats)
# library(leaps)
# library(caret)
# library(corrplot)

checkDependencies<-function(){
  GLOBAL_LIBS<-c('forcats','leaps','boot','data.table','rsq','fitdistrplus','e1071','knitr','kableExtra','corrplot','ggplot2','plyr','dplyr','devtools','caret','lmtest','mlbench')
  # automatically install any missing packages
  if (length(setdiff(GLOBAL_LIBS, rownames(installed.packages()))) > 0) {
    install.packages(setdiff(GLOBAL_LIBS, rownames(installed.packages())))
  }
  invisible(lapply(GLOBAL_LIBS,require,character.only=TRUE))
}

checkDependencies()


##########################################################
#
# Globals
##########################################################

# getStaticTempDirectory - get os platform specific temp dir that can be used
# between shessions...
getStaticTempDirectory <- function() {
  if(.Platform$OS.type =='unix')
  {
    path<-"/tmp/r-cache"
    dir.create(path, showWarnings = FALSE, recursive = TRUE, mode = "0777")
    return(path)
  }
  else if (.Platform$OS.type == 'windows')
  {
    path<-"C:/temp/r-cache"
    dir.create(path, showWarnings = FALSE, recursive = TRUE, mode = "0777")
    return(path)
  }
  else
  {
    path<-"/tmp/r-cache"
    dir.create(path, showWarnings = FALSE, recursive = TRUE, mode = "0777")
    return(path)
  }
}


CACHE_DIR <- getStaticTempDirectory()

##########################################################
#
# cached_lm:: checks local disk for a cached version of the model
#   instead of calling lm() directly, call cached_lm() and pass the same params
#   useful for consecutive r-markdown calls that will constant re-use existing models
#   jamesmb3@illinois.edu
##########################################################
cached_lm<-function(formula, data, subset, weights, na.action, method = "qr", 
    model = TRUE, x = FALSE, y = FALSE, qr = TRUE, singular.ok = TRUE, 
    contrasts = NULL, offset, rebuild= FALSE,  ...) 
{
  cl <- match.call()
  mf <- match.call(expand.dots = FALSE)
  m <- match(c("formula", "data", "subset", "weights", "na.action", "offset"), names(mf), 0L)
  fc <- mf[c(1L, m)]
  # first we build a UUID for this model using all terms 
  terms = attr(terms.formula(formula,  data = data), "term.labels")
  model_name <- paste0(terms,collapse = "+")
  # data sha
  data_sha <- digest::digest(data,"sha1")
  # add in model heuristics
  model_name <- paste0(model_name,nrow(data),data_sha)
  # digest
  model_sha <- digest::digest(model_name,"sha1")
 
  # check if file exists
  model_path <-file.path(CACHE_DIR,paste0(model_sha,".rds"))
  # if it exists, return it ...
  if(file.exists(model_path) && !rebuild ){
    print(paste("[cache] loaded model from ", model_path))
    m<-readRDS(model_path)
    return(m)
  }
  # otherwise run it, and save it ..
 else{
    #  use lm function
    fc[1]<-call(name = "lm")
    print(fc)
    m<-eval(fc)
    saveRDS(m, file = model_path)
    return(m)
  }
}

##########################################################
#
# cached_train:: checks local disk for a cached version of the model returned from caret::train() for minor adjustmests
#  for formatting, models do not have to be rediscovered
#   TODO: can be improved with additional params.
#   jamesmb3@illinois.edu
##########################################################
cached_train <-function( model, data, method, ..., weights, subset, na.action, contrasts = NULL,rebuild=FALSE)
{
  
  # first we build a UUID for this model using all terms 
  model_name <- paste0(formula(model$terms),collapse = "")
  
  # data sha
  data_sha <- digest::digest(data,"sha1")
  # add in model heuristics
  model_name <- paste0(model_name,nrow(data),data_sha)

  # digest
  model_sha <- digest::digest(model_name,"sha1")
  model_path <-file.path(CACHE_DIR,paste0(model_sha,".rds"))
  if(file.exists(model_path) && !rebuild ){
    print(paste("[cache] loaded train model from ", model_path))
    return(readRDS(model_path))
  }
  else{
    m <- train(model, data=data, method=method,...)
    saveRDS(m, file = model_path)
    return(m)
  }
  
  
}
##########################################################
#
# cached_step:: checks local disk for a cached version of the model returned from step() for aic/bic
#   instead of calling step() directly, call cached_step() and pass the same params..
#   Useful for r-markdown consecutive calls for minor adjustmests in formatting, models do not have to be rediscovered
#   TODO: can be improved with additional params.
#   jamesmb3@illinois.edu
##########################################################
cached_step <-function(object, scope, scale = 0, direction = c("both", "backward", 
    "forward"), trace = 1, keep = NULL, steps = 1000, k = 2,  rebuild=FALSE,
    ...)
{
  # build a uuid from the base model
  terms<-labels(object)
  # first we build a UUID for this model using all terms 
  model_name <- paste0(terms,collapse = "+")
   # data sha
  data_sha <- digest::digest(object$model,"sha1")
  # add in model heuristics
  model_name <- paste0(model_name,nrow(object$model),data_sha, length(coef(object)),as.character(direction) ,as.character(k))
  # digest
  model_sha <- digest::digest(model_name,"sha1")
  model_path <-file.path(CACHE_DIR,paste0(model_sha,".rds"))
  if(file.exists(model_path) && !rebuild ){
    print(paste("[cache] loaded step model from ", model_path))
    return(readRDS(model_path))
  }
  else{
    m <- step(object = object,scope = scope, scale=scale, k = k, direction = direction,trace=trace,steps=steps)
    saveRDS(m, file = model_path)
    return(m)
  }
}


##########################################################
#
# build_column_factor_map:: builds a reverse crosswalk of variable+factor -> variable
#  useful for when we are given post lm() output 
##########################################################
build_column_factor_map<-function(d){
  vs<-list()
  dfs<-lapply(names(d),function(x){
   if (is.factor(d[[x]])){
     dfs<-lapply(levels(d[[x]]),function(val){
       var <- paste0(x,val)
       return(data.frame(var=var,val=x))
     })
     return(data.table::rbindlist(dfs))
     
   }
    else
    {
       return(data.frame(var=x,val=x))
    }
  })
  return(data.table::rbindlist(dfs))
}





```

## Data Preparation

First, we load the training data and extract the number of samples and the feature names:

```{r}
train_data = read.csv('data/train.csv')
features = setdiff(colnames(train_data), c("Id", "SalePrice"))
```

The housing dataset includes **`r length(features)` features** and **`r nrow(train_data)` samples.**

### Dealing with Missing Predictor Values
Naturally, we expect that some examples are missing features, let's try removing those samples and see how much data we'd have left to work with:

```{r}
train_data_omit_na = na.omit(train_data)
nrow(train_data_omit_na)
```

Bummer! It appears every sample has is missing at least one feature. So we can't simply omit them or we'll have nothing left to work with! For factor variables with missing values, we can simply create a new *`other`* category to fill in for *`NA`* values. 

```{r}
other_category = 'other'
for (f in features) {
  if (is.factor(train_data[[f]]) && any(is.na(train_data[[f]]))) {
    levels(train_data[[f]]) = c(levels(train_data[[f]]), other_category)
    train_data[[f]][is.na(train_data[[f]])] = other_category
  }
}

# Try omitting samples with missing features again:
train_data_omit_na = na.omit(train_data)
nrow(train_data_omit_na)
```

This appears to solve most of the problem, but we're still losing `r nrow(train_data) - nrow(train_data_omit_na)` samples, since they have missing numeric feature values. Since they're a large portion (`r round((nrow(train_data)*1.0 - nrow(train_data_omit_na)) / nrow(train_data) * 100)`% of the dataset), we don't want to ignore them. Unfortunately, finding suitable replacement values for numeric features is more difficult.

Let's see which features have missing values:

```{r}
Filter(function (f) any(is.na(train_data[[f]])), features)
```

Great, we only need to deal with missing values for 3 numeric features. Let's  replace the *`NA`* values of these features with sensible defaults.

For `MasVnrArea`, the masonry veneer area in square feet, we have a large number of examples with a value of 0, so that should be a decent replacement. Since it doesn't seem especially important, we might also consider dropping the feature entirely instead of the samples.

```{r}
train_data$MasVnrArea[is.na(train_data$MasVnrArea)] = 0
```

Since `GarageYrBlt` and `YearBuilt` are highly correlated (`r cor(train_data_omit_na$GarageYrBlt, train_data_omit_na$YearBuilt)`) and they're identical `r round(sum(train_data_omit_na$GarageYrBlt == train_data_omit_na$YearBuilt) / nrow(train_data) * 100)`% of the time, we'll simply replace missing `GarageYrBlt` values with `YearBuilt` values.

```{r}
train_data$GarageYrBlt[is.na(train_data$GarageYrBlt)] = train_data$YearBuilt[is.na(train_data$GarageYrBlt)]
```

Coming up with replacement values for `LotFrontage` (linear feet of street connected to property) is much more difficult. It seems too important of a feature to simply remove. There are also no cases where the value is 0, and so no obvious replacement value.

There are a few variables like `LotArea`, `LotShape` and `LotConfig` that seem like they might be predictive of `LotFrontage`. Let's quickly build a model to predict `LotFrontage` and then use it to fill in our missing values.

```{r}
lot_frontage_model = lm(LotFrontage ~ log(LotArea) + LotArea:LotConfig + LotShape + LandSlope + YearBuilt + BldgType, data = train_data_omit_na)
```

We can be confident the `lot_frontage_model` above is doing a good job predicting `LotFrontage` because: 

* The p-value of the significance of regression test is extremely small.
* The adjusted $R^2$ is fairly large: `r summary(lot_frontage_model)$adj.r.squared`
* We're using a small number of predictors without higher degree polynomials so we aren't especially concerned with overfitting.

If we wanted to be even more confident, we might check the cross validated RMSE, but because we're only using this model to fill in missing values, we won't spend too much time on it.

Now we can fillin the missing `LotFrontage` values:

```{r}
train_data$LotFrontage[is.na(train_data$LotFrontage)] = predict(lot_frontage_model, newdata = train_data[is.na(train_data$LotFrontage),])

# Check if there are any remaining NA values in the training data:
any(is.na(train_data))
```

**Success!** We've managed to replace all of the missing values without losing any training samples.

### Creating Additional Factor Variables

When we loaded the dataset, R decided which predictors were numeric and which were factors. Some predictors that R considers numeric, might work better as factor variables. For example, the price difference between a 0 and a 2 car garage is likely different than the price difference between a 2 and a 4 car garage.

```{r}
for (f in features) {
  # We'll add any numeric predictor with less than 12 unique values as a factor variable
  if (!is.factor(train_data[[f]]) && length(unique(train_data[[f]])) <= 12) {
    train_data[[paste(f, 'Fctr', sep = "")]] = as.factor(train_data[[f]])
  }
}

# Update `features`:
features = setdiff(colnames(train_data), c("Id", "SalePrice"))
```

Great, we've added `r length(features) - 79` new factor variables for a total of **`r length(features)` features**.

### Removing Categories with Few Samples

// TODO: explain this
```{r}
category_cutoff = nrow(train_data) * 0.05
for (f in features) {
  if (is.factor(train_data[[f]])) {
    for (category in levels(train_data[[f]])) {
      samples_in_cat = sum(na.omit(train_data[[f]]) == category)
      if (samples_in_cat < category_cutoff) {
        levels(train_data[[f]]) = c(levels(train_data[[f]]), other_category)
        train_data[[f]][train_data[[f]] == category] = other_category
      }
    }
  }
}

feature_cutoff = nrow(train_data) * 0.95
for (f in features) {
  if (is.factor(train_data[[f]])) {
    if (length(levels(train_data[[f]])) == 1) {
      train_data = train_data[,!(names(train_data) == f)]
    } else {
      for (category in levels(train_data[[f]])) {
        samples_in_cat = sum(na.omit(train_data[[f]]) == category)
        if (samples_in_cat > feature_cutoff) {
          train_data = train_data[,!(names(train_data) == f)]
        }
      }
    }
  }
}
```

## Data Exploration

Next we quiclky inspect the correlations between the numeric predictors in our dataset to determine which might be useful in predicting `SalePrice` and which might have collinearity issues:

```{r fig.height=12, fig.width=14}
numeric_variables <- train_data[which(sapply(train_data, is.numeric))]
numeric_variables <- subset(numeric_variables, select = -c(Id))
correlations <- cor(numeric_variables)
corrplot(correlations, method="square", order ="FPC")
```

Next we visualize the numeric predictors that are most correlated with `SalePrice`:

```{r fig.height=12, fig.width=14}
correlated_indices = abs(correlations['SalePrice',]) > 0.2
correlated_predictors = names(correlations['SalePrice',correlated_indices])
par(mfrow = c(ceiling(length(correlated_predictors) / 5.0), 5), mar=c(1,1,1,1), oma = c(0, 0, 3, 0))
for (cor_pred in correlated_predictors) {
  plot(train_data[[cor_pred]], train_data$SalePrice, 
       main=cor_pred, col=rgb(0.1,0.1,0.1,0.3), pch=16, 
       xaxt='n', yaxt='n', xlab = NA, ylab  = NA)
}
mtext("Predictors' Relationships to SalePrice", outer=TRUE, cex = 2)
```

## Model Evaluation

As an evaluation metric, the Kaggle competition uses Root-Mean-Squared-Error (RMSE) on the *logorithm of the predicted and actual sale prices*. Taking the logorithm ensures that expensive houses don't have more of an impact on RMSE than cheap houses.

We'll use this same metric as our primary quality indicator. To ensure we don't overfit, we'll apply it on a 5-fold cross validation set. We'll also keep our eye on BIC, Adjusted $R^2$ and a few other metrics.

```{r}
get_bp_decision = function(model, alpha = 0.05) {
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_shapiro_decision = function(model, alpha = 0.05) {
  decide = unname(shapiro.test(resid(model))$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_log_rmse = function(observed, predicted) {
  sqrt(mean((log(predicted) - log(observed)) ^ 2))
}

get_cv_log_rmse = function(model_formula, data, folds = 5) {
  total_rmse = 0
  n = nrow(data)
  for (i in 1:folds) {
    idx = sample(c(TRUE, FALSE), n, replace=TRUE, prob=c(0.8, 0.2))
    training_set = data[idx,]
    cv_set = data[!idx,]
    model = cached_train(model_formula, data=training_set, method="glm")
    cv_predicted = predict(model, newdata = cv_set)
    # For stability, ignore all predictions are < 1.0
    nan_indices = is.nan(log(cv_predicted))
    total_rmse = total_rmse + get_log_rmse(cv_set$SalePrice[!nan_indices], cv_predicted[!nan_indices])
  }
  total_rmse / folds
}

evaluate_model = function(name, model, data) {
  set.seed(42)
  data.frame(Model = name,
             CV.Log.RMSE = get_cv_log_rmse(model_formula = model$terms,data =data), 
             Adj.R2 = rsq(model, adj=TRUE),
             BIC = BIC(model),
             Coefficients = length(coef(model)),
             Shapiro.Decision = get_shapiro_decision(model),
             BP.Decision = get_bp_decision(model)
             )
}
```

## Model Building


We can see that our response variable `SalePrice` has a righ skew.  A log transform can be used to adjust and normalize the distribution.

```{r echo=T}
par(mfrow=c(1,2))
hist(train_data$SalePrice,main = "SalesPrice")
hist(log(train_data$SalePrice),main = "log(SalesPrice)")
```


### Baseline Models


```{r, warning=FALSE}
# Start with a simple, naive model using all predictors added together 
additive_large = lm(SalePrice ~ ., data = train_data)
additive_large_eval = evaluate_model("Additive Large", additive_large, train_data)
kable(additive_large_eval)
```

```{r, warning=FALSE}
# Variables were selected based on the results of the Data Exploration section above
additive_small = lm(SalePrice ~ 
                      LotArea + X1stFlrSF + X2ndFlrSF + MasVnrArea + BedroomAbvGr + OverallQual + 
                      OverallCond  + ExterQual + KitchenQual + BsmtQual + BsmtExposure + ScreenPorch + 
                      Fireplaces + MSZoning + YearBuilt, data = train_data)
additive_small_eval = evaluate_model("Additive Small", additive_small,train_data)
kable(additive_small_eval)
```

We can do much better on our target metric (CV.Log.RMSE) with the smaller model. Notice that the small additive model has far fewer parameters than the large one (`r length(coef(additive_small))` vs  `r length(coef(additive_large))`) It seems that with so many predictors, it's very easy to overfit to the training data. 

### Parameter Search using BIC

The small baseline model also has a better BIC. So it looks like BIC might be a good metric to use to avoid overfitting. Next, we'll use BIC to do a backwards parameter search starting from `additive_large` model.

```{r, warning=FALSE}
additive_bic_backward = cached_step(additive_large, k = log(nrow(train_data)), direction = 'backward', trace = 0)
additive_bic_backward_eval = evaluate_model("Additive Backward BIC", additive_bic_backward,train_data)
kable(additive_bic_backward_eval)
```

### Transformations
Doing a parameter search using backwards BIC managed improve our results, but it still fails the Shapiro-Wilks and Breusch-Pagan tests. This suggests we might be able to achieve better results by applying some transformations to our the response and/or predictors.

Based on the graphs in the *Data Exploration* section above, it appears that many predictors have an exponential or polynomial relationship with `SalePrice`.

# TODO: log transform etc...



# Log Transform Continuous Variables
```{r}

numeric_variables <- names(which(sapply(train_data, is.numeric)))
categorical_variables <- setdiff(names(train_data),(numeric_variables))
# remove Id
numeric_variables <- numeric_variables[numeric_variables!= "Id" & numeric_variables!= "SalePrice"]
numeric_variables_data <- subset(train_data[numeric_variables])

par(mfrow=c(2,2))
tmp<-sapply(numeric_variables,function(var){
  hist(train_data[[var]],main = var)
  hist(log(train_data[[var]]),main = paste0("log(",var,")"))
})

should_log_transform<-c('MSSubClass','LotArea','BsmtFinSF1','BsmtFinSF2','BsmtUnfSF','TotalBsmtSF','X1stFlrSF','X2ndFlrSF','LowQualFinSF','GrLivArea','BedroomAbvGr','TotRmsAbvGrd','GarageArea','WoodDeckSF','OpenPorchSF','EnclosedPorch','X3SsnPorch','ScreenPorch','PoolArea','MiscVal','MoSold')

```


```{r}

new_train_data<-train_data
# apply log transforms to predictors
tmp<-sapply(should_log_transform,function(var){
  new_train_data[[var]] <<- log(1+ new_train_data[[var]]  )
})



all_factor_column_names<-build_column_factor_map(d = new_train_data)
model_additive = lm(log(SalePrice) ~ . -Id-MSSubClass, data = new_train_data)
pvals<-coef(summary(model_additive))[,"Pr(>|t|)"]
# get pvals less than .001 and make a new formula
pval_cols<-data.frame(var=names(pvals[pvals<0.001]))
pval_cols_used<-unique(merge(all_factor_column_names,pval_cols,by.x = "var",by.y = "var")$val)
pval_formula<-as.formula(paste("log(SalePrice)~", paste(pval_cols_used, collapse="+")))

influential_indexes <- as.vector(which(cooks.distance(model_additive) > (4 / length(cooks.distance(model_additive)))))
influential_indexes <- c(influential_indexes,as.vector(which(abs(resid(model_additive))>.5,0)))
pval_model<-lm(pval_formula, data=new_train_data[-influential_indexes])
pval_model_eval<-evaluate_model("pval",pval_model,new_train_data[-influential_indexes])

p <- predict(pval_model, newdata = new_train_data[-influential_indexes])
plot(exp(p), 
     new_train_data$SalePrice, 
     col = "dodgerblue", 
     pch = 20,
     main = "Prediction vs Actual",
     xlab = "Prediction",
     ylab = "Actual"
     )

```
# Results

```{r}
model_results = rbind(additive_large_eval, additive_small_eval, additive_bic_backward_eval,pval_model_eval)
kable(model_results)
```

```{r}
best_model = additive_bic_backward # TODO: replace with model that's actually the best
```

```{r}
# Fitted vs. Residuals Plot
plot(fitted(best_model), resid(best_model), col = 'grey',
     xlab = "Fitted", ylab = "Residuals", main = "Fitted vs. Residuals")
    abline(h = 0, col = 'orange', lwd = 2)
```

```{r}
# Q-Q Plot
par(mfrow=(c(1,1)))
qqnorm(resid(pval_model), main = "Normal Q-Q Plot", col = 'grey')
qqline(resid(pval_model), col = 'dodgerblue', lwd = 2)
```

# Discussion


Lets summarize what we did, we started with a dataset with a number of invalid values and a large number of predictors. Then we started with the largest possible additive model and reduced number of predictors using backward selection and AIC. We tested resulting model for LINE assumptions for linear regression and found that those assumptions are violated.
Then we derived another model using predictor and response tranformation and train it on a dataset that excludes infuential data points. We tested this new model for LINE assumptions and found it was in line with these assumptions, finally we used test dataset for prediction and calculated train and test RMSE. Of course this model is still not the best possible model but can be used for some prediction and can be further iproved by applying further transformations.
The final selected model uses both numerical and categorical predictors. Some of the notable predictors that become part of selected model are overall quality, year bulit, ground living area, overall contidions, garage area,lot area etc.. Clearly these all play a major role for determining sales price of a home. So we can conclude that selected model is in line with the practical considerations of housing market.
We also tried using interactive predictiors but with such a large number of categorial predictors, the number of predictors were going very high and keeping R studio busy for hours. 
In future we would like to run R studio in more powerfull machine and explore areas of multithreaded processing to speed up training the model and submit result on Kaggle competition to see the result.


We have spent many hours fitting various models, and most models did well with Adjusted-$$R^2$$ near 90% which means that most of them behave reasonable well - we would prefer models with less coefficients because they are easier to explain - however we noticed there is a balance between including too many coefficients that tend to overfit vs very few coefficients that are hard to explain.


#### Challenges
The dataset we chose was challenging for a variety of reasons, we think that because there are so many columns and so few rows we can easily run into overfitting - so perhaps in the future we would be able to expand the number of observations.
We have also seen that some of the AIC/BIC/train functions may take extremely long times - so `caching` was added so that we would be able to perform iterative updates to the formatting and text.

#### Future Work


# Acknowledgments

We would like to thank [Professor David Unger](https://stat.illinois.edu/directory/profile/dunger) for his input has been invaluable during the course and lecture hours and also [Professor David Dalpiaz](http://daviddalpiaz.github.io/appliedstats/) for the material that was covered in the course.

